---
title: "Reproductive maturity analysis results of Patella spp. in Madeira"
author: "Eros Fernando Geppi"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Data manipulation
library(tidyverse)   # Includes dplyr, ggplot2, tibble, readr, etc.
library(lubridate)   # Date and time parsing
library(janitor)     # Data cleaning
library(readxl)      # Excel file import
library(stringr)     # String manipulation
library(tidyr)       # Data reshaping
library(dplyr)
# Visualization
library(ggplot2)     # Base plotting
library(scales)      # Axis formatting
library(cowplot)     # Plot composition
library(ggthemes)    # Custom themes
library(hrbrthemes)  # Additional themes
library(viridis)     # Color palettes
library(ggridges)    # Ridgeline plots
library(ggrepel)     # Smart text labels
library(gridExtra)   # Grid layouts
library(grid)        # Base grid system
library(ggpubr)      # Publication-ready plots

# Spatial data / maps
library(sf)              # Simple Features for spatial data
library(ggspatial)       # Scale bars and north arrows
library(ggmap)           # Map backgrounds
library(geodata)         # GADM shapefiles
library(rnaturalearth)   # Country outlines
library(rnaturalearthdata)

# Statistical modeling
library(broom)       # Tidy model output
library(emmeans)     # Estimated marginal means
library(car)         # ANOVA, VIF
library(multcomp)    # Multiple comparisons
library(lmtest)      # Model testing
library(MASS)        # Classic datasets and models
library(moderndive)  # Teaching statistics with tidy tools

# Statistical tests & summaries
library(moments)     # Skewness, kurtosis
library(DescTools)   # Descriptive statistics
library(descr)       # Frequency tables
library(effsize)     # Effect sizes
library(vcd)         # Visualizing categorical data
library(conover.test)# Conover test (non-parametric)
library(pROC)        # ROC curves
library(flextable)
```

## 0. Imports and cleaning data

```{r}
Mad_RAW <- readxl::read_excel("E:/Investigación/Repositorios/Lapacom/Data/ToAnalyze/Madeira/BD_LIMPETS_MAD_1996-2018.xlsx", sheet = "Data", range = "A1:T69707")
Mad_RAW_Clean <- Mad_RAW %>% clean_names()

Mad_selected <- Mad_RAW_Clean %>%
  dplyr::select(species, year, month, total_length_mm, total_length_class_mm, weight_g, mature_imature,
         sampling_site, lat, long, protective_regime, proximity_human_settlements,
         accessibility, age_lt, age_months, age_class)

Mad_2sp <- Mad_selected %>%
  filter(species %in% c("Patella ordinaria", "Patella aspera")) %>%
  filter(!is.na(sampling_site) & trimws(sampling_site) != "") %>%
  mutate(across(where(is.character), as.factor),
         regulation_period = if_else(year < 2007, "Before", "After") %>% as.factor())

Mad_2sp_analisis <- Mad_2sp %>%
  filter(sampling_site %in% c("Porto Moniz", "Paúl do Mar", "Funchal", "Desertas",
                              "Caniçal", "Santa Cruz", "Ribeira Brava", "São Vicente")) %>%
  mutate(
    across(where(is.character), as.factor),
    regulation_period = factor(regulation_period, levels = c("Before", "After"))
  )

Mad_2sp_analisis <- Mad_2sp_analisis %>%
  mutate(
    accessibility = case_when(
      protective_regime == "MPA" ~ "South",              # asignar categoría South a todas las MPA
      TRUE ~ as.character(accessibility)                 # mantener el resto tal como está
    )
  ) %>%
  filter(!is.na(accessibility), accessibility %in% c("North", "South")) %>%  # eliminar NA y "North-South"
  mutate(
    accessibility = factor(accessibility, levels = c("North", "South"))      # asegurar orden y formato
  )
summary(Mad_2sp_analisis)

```

## 1. Localización geográfica de los sitios de muestreo

```{r mapa-muestreo}

# shapefiles and polígonos

world <- ne_countries(scale = 10, returnclass = "sf")
portugal_gadm <- geodata::gadm("PRT", level = 1, path = tempdir()) %>% st_as_sf()
portugal_mainland <- portugal_gadm %>% filter(!NAME_1 %in% c("Madeira", "Açores"))
madeira_sf <- portugal_gadm %>% filter(NAME_1 == "Madeira")
azores_sf <- portugal_gadm %>% filter(NAME_1 == "Açores")

# Coordenadas DMS a decimales
convert_dms_to_decimal <- function(dms) {
  parts <- str_match(dms, "(\\d+)°(\\d+)'(\\d+\\.?\\d*)\"?([NSEW])")
  deg <- as.numeric(parts[, 2])
  min <- as.numeric(parts[, 3])
  sec <- as.numeric(parts[, 4])
  dir <- parts[, 5]
  decimal <- deg + min / 60 + sec / 3600
  ifelse(dir %in% c("S", "W"), -decimal, decimal)
}

coords <- Mad_2sp_analisis %>%
  dplyr::select(sampling_site, lat, long,protective_regime) %>%
  distinct() %>%
  drop_na() %>%
  mutate(
    lat_dd = convert_dms_to_decimal(lat),
    long_dd = convert_dms_to_decimal(long)
  )

coords_sf <- st_as_sf(coords, coords = c("long_dd", "lat_dd"), crs = 4326)

etiquetas <- tibble::tibble(
  lugar = c("Portugal", "Selvagens", "Azores", "Canary Islands", "Morocco", "Gibraltar", "Spain"),
  lon = c(-10.5, -16.1, -26.0, -16.5, -8.5, -5.4, -6.5),
  lat = c(39.5, 31.0, 38.6, 29.2, 31.0, 36.1, 43.0)
)

# Mapas
map_region <- ggplot() +
  geom_sf(data = world, fill = "grey90", color = "darkgrey") +
  geom_sf(data = portugal_mainland, fill = "darkgrey") +
  geom_sf(data = madeira_sf, fill = "darkgrey") +
  geom_sf(data = azores_sf, fill = "darkgrey") +
  geom_text(data = etiquetas, aes(x = lon, y = lat, label = lugar),
            size = 3.2, fontface = "italic") +
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotate("rect", xmin = -17.6, xmax = -16.2, ymin = 32.4, ymax = 33.2,
           fill = "grey", alpha = 0.3, color = "grey20", size = 0.7) +
  annotate("text", x = -20, y = 36, label = "Atlantic Ocean", angle = 45, size = 7,
           fontface = "italic", color = "gray30") +
  annotate("text", x = -17.0, y = 33.5, label = "Madeira", size = 3.2, fontface = "bold",
           color = "black") +
  annotation_north_arrow(location = "tl", which_north = "true",
                         style = north_arrow_fancy_orienteering()) +
  coord_sf(xlim = c(-30, -5), ylim = c(27, 44), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude") +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(color = "darkgrey", size = 0.2),
    axis.title = element_blank(),
    plot.margin = margin(5, 5, 5, 5)
  )

map_madeira <- ggplot() +
  geom_sf(data = madeira_sf, fill = "gray90", color = "darkgrey") +
  geom_sf(data = coords_sf, aes(color = sampling_site, shape=protective_regime), size = 3) +
  coord_sf(xlim = c(-17.6, -16.2), ylim = c(32.4, 33.2), expand = FALSE) +
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotation_north_arrow(location = "tl", which_north = "true",
                         style = north_arrow_fancy_orienteering()) +
  labs(color = "Sitios de muestreo") +
  theme_minimal() +
  theme(legend.position = "bottom")

legend_b <- ggplotGrob(map_madeira) %>%
  gtable::gtable_filter("guide-box")
map_madeira_noleg <- map_madeira + theme(legend.position = "none")

panel_maps <- plot_grid(
  map_region, map_madeira_noleg,
  labels = c("A)", "B)"),
  label_size = 14,
  ncol = 2,
  rel_widths = c(1, 1.2)
)

final_plot <- plot_grid(
  panel_maps,
  legend_b,
  ncol = 1,
  rel_heights = c(1, 0.1)
)

final_plot
```

```{r}

madurez_prop_mensual_pr <- Mad_2sp_analisis %>%
  filter(!is.na(mature_imature)) %>%
  group_by(year, month, species, sampling_site, accessibility, protective_regime) %>%
  summarise(
    total = n(),
    maduros = sum(mature_imature == "Mature"),
    proporcion_madura = maduros / total,
    .groups = "drop"
  ) %>%
  mutate(
    grupo = interaction(species, protective_regime, sep = " - "),
    fecha = lubridate::make_date(year, month, 1)  # For a continuous time axis
  )

# Visual style for protection regime plot
colores <- c(
  "Patella aspera - Full acess" = "#D39C00",
  "Patella ordinaria - Full acess" = "#0072B2",
  "Patella aspera - MPA" = "#D39C00",
  "Patella ordinaria - MPA" = "#0072B2"
)

lineas <- c(
  "Patella aspera - Full acess" = "solid",
  "Patella ordinaria - Full acess" = "solid",
  "Patella aspera - MPA" = "dotted",
  "Patella ordinaria - MPA" = "dotted"
)

# Final plot: mature proportion by protection regime
MATURE_PROP_protreg <- ggplot(
  data = madurez_prop_mensual_pr, 
         aes(x = fecha,
             y = proporcion_madura,
             color = grupo,
             linetype = grupo)) +
  
  geom_line(linewidth = 1) +
  
  geom_vline(
    xintercept = as.Date("2007-01-01"),
    linetype = "dashed",
     color = "black",
     linewidth = 0.7) +
  
  scale_color_manual(
    values = colores) +
  
  scale_linetype_manual(
    values = lineas) +
  
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1)) +
  
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_date(
    date_breaks = "2 year",
    date_labels = "%Y"
  ) +
  
  labs(
    x = "Date",
    y = "Mature Proportion (%)") +
  
  theme_minimal() +
  
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )

# Monthly proportion of mature individuals by accessibility
madurez_prop_mensual_ac <- Mad_2sp_analisis %>%
  filter(!is.na(mature_imature)) %>%
  group_by(year, month, species, sampling_site, accessibility, protective_regime) %>%
  summarise(
    total = n(),
    maduros = sum(mature_imature == "Mature"),
    proporcion_madura = maduros / total,
    .groups = "drop"
  ) %>%
  mutate(
    grupo = interaction(species, accessibility, sep = " - "),
    fecha = lubridate::make_date(year, month, 1)  # For a continuous time axis
  )

# Visual style for accessibility plot
colores <- c(
  "Patella aspera - North" = "#D39C00",
  "Patella ordinaria - North" = "#0072B2",
  "Patella aspera - South" = "#D39C00",
  "Patella ordinaria - South" = "#0072B2"
)

lineas <- c(
  "Patella aspera - South" = "solid",
  "Patella ordinaria - South" = "solid",
  "Patella aspera - North" = "dotted",
  "Patella ordinaria - North" = "dotted"
)

# Final plot: mature proportion by accessibility
MATURE_PROP_ACC <- ggplot(
  madurez_prop_mensual_ac,
  aes(x = fecha,
      y = proporcion_madura,
      color = grupo, 
      linetype = grupo)) +
  
  geom_line(
    linewidth = 1) +
  
  geom_vline(
    xintercept = as.Date("2007-01-01"),
    linetype = "dashed",
    color = "black",
    linewidth = 0.7) +
  
  scale_color_manual(
    values = colores) +
  
  scale_linetype_manual(
    values = lineas) +
  
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1)) +
  
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_date(
    date_breaks = "2 year",
    date_labels = "%Y"
  ) +
  
  labs(
    x = "Date",
    y = "Mature proportion (%)") +
  
  theme_minimal() +
  
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )

# Plot both panels (A and B)
plot_grid(
  MATURE_PROP_protreg, MATURE_PROP_ACC,
  labels = c("A)", "B)"),
  label_size = 16,
  label_x = 0.02,   # Horizontal alignment (closer to edge)
  label_y = 0.3,    # Vertical alignment
  hjust = 0,        # Left justified
  vjust = 1,        # Top justified
  ncol = 1,
  rel_heights = c(1, 1)
)
```

## 2. Monthly proportion of mature adults per year.

```{r}
madurez_prop <- Mad_2sp_analisis %>%
  filter(!is.na(mature_imature)) %>%
  group_by(year, month, species, protective_regime, regulation_period) %>%
  summarise(
    total = n(),
    maduros = sum(mature_imature == "Mature"),
    prop_maduros = maduros / total,
    .groups = "drop"
  ) %>%
  mutate(fecha = make_date(year, month, 1))

ggplot(madurez_prop, 
       aes(x = fecha, y = prop_maduros, 
           color = interaction(protective_regime, regulation_period), linetype = species)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = as.Date("2007-01-01"), 
             linetype = "dashed", color = "black") +
  labs(
    title = "Monthly evolution of the proportion of mature individuals",
    x = "Year",
    y = "Mature proportion (%)",
    color = "Protection × Period",
    linetype = "Species"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_date(
  date_breaks = "2 year",
  date_labels = "%Y"
  )+
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

## 3. Comparación antes y después de 2007

```{r}
madurez_yearly <- madurez_prop %>%
  group_by(year, species, protective_regime, regulation_period) %>%
  summarise(prop_m = mean(prop_maduros, na.rm = TRUE), .groups = "drop")

ggplot(madurez_yearly, aes(x = regulation_period, y = prop_m, fill = protective_regime)) +
  geom_boxplot() +
  facet_wrap(~species) +
  labs(title = "Mature adult proportion befor and after 2007", y = "Mature Proportion (%)", x = "Regulation") +
  theme_minimal()
```

## 4. Estacionalidad y periodos de veda

```{r}
madurez_mes <- madurez_prop %>%
  group_by(month, species, protective_regime) %>%
  summarise(prop_m = mean(prop_maduros, na.rm = TRUE), .groups = "drop")

periodos <- tibble(
  month = 1:12,
  periodo = case_when(
    month %in% 4:9 ~ "Exploitation",
    TRUE ~ "Reproductive"
  )
)

madurez_mes_plot <- madurez_mes %>%
  left_join(periodos, by = "month")

ggplot(madurez_mes_plot, aes(x = month, y = prop_m, 
                             color = protective_regime, 
                             linetype = species)) +
  # Fondo por periodo con baja opacidad
  geom_tile(aes(x = month, fill = periodo, y = 0.5), height = 1, alpha = 0.1,
            inherit.aes = FALSE, show.legend = FALSE) +

  # Línea y puntos
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +

  # Escalas
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = c("Exploitation" = "mistyrose", 
                               "Reproductive" = "palegreen")) +

  # Etiquetas
  labs(title = "Seasonality of maturity and exploitation vs reproductive period",
       x = "Month", y = "Mature proportion") +

  # Tema limpio
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.box = "horizontal",
    legend.position = "bottom",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

```

## 5. Modelo GLM binomial para evaluar efecto MPA y regulación

### Patella aspera

```{r}
# Modelo sobre datos individuales para Patella aspera
modelo_comparado_pa <- Mad_2sp_analisis %>%
filter(species == "Patella aspera", !is.na(mature_imature)) %>%
mutate(
maduro = mature_imature == "Mature",
grupo = case_when(
protective_regime == "Full acess" & regulation_period == "Before" ~ "FA_Before",
protective_regime == "Full acess" & regulation_period == "After" ~ "FA_After",
protective_regime == "MPA" & regulation_period == "After" ~ "MPA_After"
),
grupo = factor(grupo, levels = c("FA_Before", "FA_After", "MPA_After"))
) %>%
glm(maduro ~ grupo, data = ., family = binomial())
# Resultados del modelo (odds ratios con IC 95%)
broom::tidy(modelo_comparado_pa, exponentiate = TRUE, conf.int = TRUE)

em_pa <- emmeans(modelo_comparado_pa, ~ grupo, type = "response")

contrastes_interes_pa <- contrast(em_pa, method = list(
  "FA_Before vs FA_After" = c(1, -1, 0),
  "FA_Before vs MPA_After" = c(1, 0, -1),
  "FA_After vs MPA_After" = c(0, 1, -1)
), type = "response")


# Convertir resultados de emmeans a data frame
contrastes_df_pa <- as.data.frame(summary(contrastes_interes_pa))

# Formatear columnas:
# - Redondear todas las columnas numéricas excepto p.value
# - Formatear p.value en notación científica
tabla_formateada_pa <- contrastes_df_pa %>%
  mutate(
    across(where(is.numeric) & !matches("p.value"), ~ round(., 3)),
    p.value = format(p.value, scientific = TRUE, digits = 6)
  )

# Mostrar como flextable con título
flextable(tabla_formateada_pa) %>%
  set_caption(caption = "Marginal Effects (EMMEANS) - Patella aspera")
```

### Patella ordinaria

```{r}
modelo_comparado_po <- Mad_2sp_analisis %>%
filter(species == "Patella ordinaria", !is.na(mature_imature)) %>%
mutate(
maduro = mature_imature == "Mature",
grupo = case_when(
protective_regime == "Full acess" & regulation_period == "Before" ~ "FA_Before",
protective_regime == "Full acess" & regulation_period == "After" ~ "FA_After",
protective_regime == "MPA" & regulation_period == "After" ~ "MPA_After"
),
grupo = factor(grupo, levels = c("FA_Before", "FA_After", "MPA_After"))
) %>%
glm(maduro ~ grupo, data = ., family = binomial())
# Resultados del modelo (odds ratios con IC 95%)
broom::tidy(modelo_comparado_po, exponentiate = TRUE, conf.int = TRUE)

em_po <- emmeans(modelo_comparado_po, ~ grupo, type = "response")

contrastes_interes_po <- contrast(em_po, method = list(
  "FA_Before vs FA_After" = c(1, -1, 0),
  "FA_Before vs MPA_After" = c(1, 0, -1),
  "FA_After vs MPA_After" = c(0, 1, -1)
), type = "response")

em_po <- emmeans(modelo_comparado_po, ~ grupo, type = "response")

contrastes_interes_po <- contrast(em_po, method = list(
  "FA_Before vs FA_After" = c(1, -1, 0),
  "FA_Before vs MPA_After" = c(1, 0, -1),
  "FA_After vs MPA_After" = c(0, 1, -1)
), type = "response")


# Convertir resultados de emmeans a data frame
contrastes_df_po <- as.data.frame(summary(contrastes_interes_po))

# Formatear columnas:
# - Redondear todas las columnas numéricas excepto p.value
# - Formatear p.value en notación científica
tabla_formateada_po <- contrastes_df_po %>%
  mutate(
    across(where(is.numeric) & !matches("p.value"), ~ round(., 3)),
    p.value = format(p.value, scientific = TRUE, digits = 6)
  )

# Mostrar como flextable con título
flextable(tabla_formateada_po) %>%
  set_caption(caption = "Marginal Effects (EMMEANS) - Patella ordinaria")
```

## 6. Gráfico de odds ratios del modelo de implementacion de MPA y regulaciones.

```{r}
# Obtener resultados tidy para ambas especies
po_df <- broom::tidy(modelo_comparado_po, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = dplyr::recode(term,
                         "grupoFA_After" = "Full access (After)",
                         "grupoMPA_After" = "MPA (After)"),
    species = "Patella ordinaria"
  )

pa_df <- broom::tidy(modelo_comparado_pa, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = dplyr::recode(term,
                         "grupoFA_After" = "Full access (After)",
                         "grupoMPA_After" = "MPA (After)"),
    species = "Patella aspera"
  )
# Combinar en un solo data frame
odds_df <- bind_rows(po_df, pa_df)

# Graficar
ggplot(odds_df, aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high, color = species)) +
  geom_pointrange(position = position_dodge(width = 0.5), size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
  coord_flip() +
  labs(
    title = "Odds ratios by species and management regime",
    x = "", y = "Odds Ratio (95% CI)", color = "Species"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom"
  )

```

## 7. Efecto de la accesibilidad (conectividad)

Evaluate the effect of geographic accessibility (North/South) on the proportion of mature individuals, considering the interaction with:

protective regime (Total access / MPA)

regulatory period (Before / After)

### Efecto de la accesibilidad en zonas Full access – Patella ordinaria

```{r}

resumen_acc_po <- Mad_2sp_analisis %>%
  filter(
    species == "Patella ordinaria",
    protective_regime == "Full acess",
    !is.na(mature_imature),
    !is.na(accessibility)
  ) %>%
  mutate(
    maduro = mature_imature == "Mature"
  ) %>%
  group_by(accessibility) %>%
  summarise(
    maduros = sum(maduro),
    total = n(),
    .groups = "drop"
  ) %>%
  mutate(
    accessibility = factor(accessibility, levels = c("North", "South"))
  )

modelo_acc_po <- glm(
  cbind(maduros, total - maduros) ~ accessibility,
  data = resumen_acc_po,
  family = binomial()
)

broom::tidy(modelo_acc_po, exponentiate = TRUE, conf.int = TRUE)
```

### Efecto de accesibilidad en zonas Full access -- Patella aspera

```{r}

resumen_acc_pa <- Mad_2sp_analisis %>%
  filter(
    species == "Patella aspera",
    protective_regime == "Full acess",
    !is.na(mature_imature),
    !is.na(accessibility)
  ) %>%
  mutate(
    maduro = mature_imature == "Mature"
  ) %>%
  group_by(accessibility) %>%
  summarise(
    maduros = sum(maduro),
    total = n(),
    .groups = "drop"
  ) %>%
  mutate(
    accessibility = factor(accessibility, levels = c("North", "South"))
  )

modelo_acc_pa <- glm(
  cbind(maduros, total - maduros) ~ accessibility,
  data = resumen_acc_pa,
  family = binomial()
)

broom::tidy(modelo_acc_pa, exponentiate = TRUE, conf.int = TRUE)
```

## 8. Gráfico de odds ratios de modelos de accesibilidad

```{r accesibilidad-lineas, fig.width=8, fig.height=5}
# Patella ordinaria
res_po <- broom::tidy(modelo_acc_po, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(species = "Patella ordinaria")

# Patella aspera
res_pa <- broom::tidy(modelo_acc_pa, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(species = "Patella aspera")

# Unimos ambos modelos
res_combinado <- bind_rows(res_po, res_pa) %>%
  dplyr::mutate(term = dplyr::recode(term,
    "accessibilitySouth" = "South vs North"
  ))

# Gráfico
ggplot(res_combinado, aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high, color = species)) +
  geom_pointrange(position = position_dodge(width = 0.6), size = 0.9) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  coord_flip() +
  labs(
    title = "Odds Ratios – Efecto de accesibilidad (Full access)",
    x = "", y = "Odds Ratio (IC 95%)", color = "Especie"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom"
  )
```
